---
title: "World Health Organization (WHO) Cholesterol Data vs. Cardiovascular Death Rates"
author: "Steve Bargelt"
date: "August 31, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(cache=TRUE)
```

### Load packages

```{r load-packages, message=FALSE}
library(WHO)
library(plyr)
library(dplyr)
library(ggplot2)
library(scales)
library(showtext)
library(scatterD3)
library(stringr)
library(RColorBrewer)
library(dygraphs)
```

## Background and Inspiration

The investigation will focus on World Health Organization data on cholesterol data and correlation to cardiovascular (CVD) as casue of death. 

### Sources and More Information

You can find all of the source code at [Github](https://github.com/stevebargelt/WHO-Data "WHO Cholesterol Data").

[World Health Organization (WHO) Global Health Observatory (GHO)](http://www.who.int/gho/en/)

[WHO Cholesterol Data](http://apps.who.int/gho/data/node.main.A883?lang=en)

[WHO Ischaemic Data](https://rho.emro.who.int/rhodata/node.main.A1079?lang=en)

[WHO Cerebrovascular Data](https://rho.emro.who.int/rhodata/node.main.A1073?lang=en)

## Data Acquisition

The data takes quite a while to download from the WHO (3-4 minutes on a fast internet connection). You will find a stand-alone R script in the Github repo that will save the data to local files to speed up ad-hoc analysis.

### Use the WHO API to get the data we want to analyze.
```{r download-data}

CHO_Age_Stand <- get_data("CHOL_03")
CHO_Crude <- get_data("CHOL_04")
#CVD_Cerebrovascular_DALY <- get_data("SA_0000001689")
CVD_Cerebrovascular <- get_data("SA_0000001690")
#CVD_Ischaemic_DALY <- get_data("SA_0000001425")
CVD_Ischaemic <- get_data("SA_0000001444")

```

## Tidying up the data

The Cholesterol (CHO) values are stored in the as a character string "5.0 [4.8-5.3]" - the following code strips out everything between and including the [ ] leaving us with a numeric value for each entry.
```{r clean-cho-data}
CHO_Age_Stand$value <- gsub( " *\\[.*?\\] *", "", CHO_Age_Stand$value)
CHO_Age_Stand$value <- as.numeric(CHO_Age_Stand$value)

CHO_Crude$value <- gsub( " *\\[.*?\\] *", "", CHO_Crude$value)
CHO_Crude$value <- as.numeric(CHO_Crude$value)
```

Combine [Cerebrovascular](https://en.wikipedia.org/wiki/Cerebrovascular_disease) and [Ischaemic](https://en.wikipedia.org/wiki/Coronary_artery_disease) death rates to get a picture of all Cardiovascualr death rates.
```{r }
CVD_ALL <- merge(CVD_Cerebrovascular, CVD_Ischaemic, by=c("year", "region", "country", "sex", "publishstate"))
CVD_ALL$totalCVDvalue = CVD_ALL$value.x + CVD_ALL$value.y
colnames(CVD_ALL)[7] <- "cerrbrovascularValue"
colnames(CVD_ALL)[9] <- "ischaemicValue"
CVD_ALL$gho.y <- NULL
CVD_ALL$gho.x <- NULL
```

Merging cholesterol data with all cardiovascular death data
```{r }
db <- merge(CVD_ALL, CHO_Crude, by=c("year", "region", "country", "sex", "publishstate"))
colnames(db)[11] <- "choCrudeValue"
db$gho <- NULL

choDeathData <- merge(db, CHO_Age_Stand, by=c("year", "region", "country", "sex", "publishstate", "agegroup"))
colnames(choDeathData)[12] <- "choAgeStandValue"
choDeathData$gho <- NULL
choDeathData$publishstate <- NULL
choDeathData$agegroup <- NULL
choDeathData$sex <- as.factor((choDeathData$sex))

```

### Age Standardized vs. Crude Cholesterol Numbers

Must decide on one or the other since not all values are present for all data points. I have decided to use Age Standardized value since there are ~20 more complete cases.

Mean Age Standardized: `r mean(choDeathData$choAgeStandValue, na.rm = TRUE)` <br />
Mean Crude: `r mean(choDeathData$choCrudeValue, na.rm = TRUE)` <br />
Median Age Standardized: `r median(choDeathData$choAgeStandValue, na.rm = TRUE)` <br />
Median Crude: `r median(choDeathData$choCrudeValue, na.rm = TRUE)` <br />
IQR Age Standardized: `r IQR(choDeathData$choAgeStandValue, na.rm = TRUE)` <br />
IQR Crude: `r IQR(choDeathData$choCrudeValue, na.rm = TRUE)` <br />
Number of complete samples Crude: `r  nrow(choDeathData[complete.cases(choDeathData[,c(7,8)]),])` <br />
Number of complete samples Age Standardized: `r nrow(choDeathData[complete.cases(choDeathData[,c(7,9)]),])` <br />

```{r remove choAgeStanValue}
#choDeathData$choCrudeValue <- NULL
choDeathData <- choDeathData[complete.cases(choDeathData[,c(7,8)]),]

```

## Analysis

```{r CHOcrude, echo=FALSE}
r2<-ddply(choDeathData,.(sex),function(x) summary(lm(x$totalCVDvalue ~ x$choAgeStandValue))$r.squared)
names(r2)<-c("sex","r2")

r<-ddply(choDeathData,.(sex),function(x) cor(x$totalCVDvalue, x$choAgeStandValue, , method="pearson"))
names(r)<-c("sex","r")

g<-ggplot(choDeathData, aes(x=choAgeStandValue,y=totalCVDvalue,group=sex))
g<-g+geom_point(aes(color=sex))
g<-g+geom_text(data=r2,aes(color=sex, label = paste("R^2: ", r2,sep="")),parse=T,x=5.5,y=c(800,750), show.legend=TRUE)
g<-g+geom_text(data=r,aes(color=sex, label = paste("R: ", r,sep="")),parse=T,x=5.5,y=c(700,650), show.legend=TRUE)
g<-g+theme_light(9)
g+labs(x = "Age standardized cholesterol Mean mmol/L\n", y = "CVD Death rate per 100,000", 
       linetype = NULL, color = NULL,
       title = "Cardiovascualr Death Rate vs. Mean Cholesteral (Age Standardized) (by region)\n")
```



This chart is interactie. You can zoom in and out on the data points. You can also hover over the Legend on the right to highlight only the data that corresponds to the symbol you are hovering over. Hovering over a datapoint give your more information. Finally you can lasso (SHIFT-Click and Drag) a set of points to single them out.

```{r CHO_death_interactive, echo=FALSE}

tooltips <- paste("<strong>", choDeathData$country,"</strong><br />", 
                  "Region: ", choDeathData$region, "<br />",
                  "CHO: ", choDeathData$choAgeStandValue, "<br />",
                  "Deaths: ", choDeathData$totalCVDvalue
                  )

scatterD3(x = choDeathData$choAgeStandValue, 
          y = choDeathData$totalCVDvalue, 
          col_var=choDeathData$sex,
          symbol_var = choDeathData$region,
          xlab = "Cholesterol mmol//L (Age Standardized)", 
          ylab = "Death rate per 100,000 from all CVD", 
          col_lab = "Gender",
          symbol_lab = "Regions",
          lasso = TRUE, 
          ellipses = TRUE,
          tooltip_text = tooltips)
```


```{r maleonly, echo=FALSE}

choDeathData %>% 
  filter(sex == "Male") %>% 
  ggplot(aes(x=choCrudeValue, y=totalCVDvalue)) +
  geom_point(shape=1) +    # Use hollow circles
  geom_smooth(method = lm) +
  theme_light(9) +
  labs(x = "Cholesterol Mean mmol/L\n", y = "Male CVD Death rate per 100,000", 
       linetype = NULL, color = NULL,
       title = "Male Cardiovascualr Death Rate vs. Mean Cholesteral (crude) (by region)\n")

```


```{r femaleCHOCrude, echo=FALSE}
choDeathData %>% 
  filter(sex == "Female") %>% 
  ggplot(aes(x=choCrudeValue, y=totalCVDvalue)) +
  geom_point(shape=1) +    # Use hollow circles
  geom_smooth(method = lm) +
  theme_light(9) +
  labs(x = "Cholesterol Mean mmol/L\n", y = "Female CVD Death rate per 100,000", 
       linetype = NULL, color = NULL,
       title = "Female Cardiovascular Death Rate vs. Mean Cholesteral (crude) (by region)\n")

```


```{r maleCHOAgeStand, echo=FALSE}
choDeathData %>% 
  filter(sex == "Male") %>% 
  ggplot(aes(x=choAgeStandValue, y=totalCVDvalue)) +
  geom_point(shape=1) +    # Use hollow circles
  geom_smooth(method = lm) +
  theme_light(9) +
  labs(x = "Cholesterol Mean mmol/L\n", y = "Male CVD Death rate per 100,000", 
       linetype = NULL, color = NULL,
       title = "Male Cardiovascualr Death Rate vs. Mean Cholesteral (Age Standardized) (by region)\n")

```


```{r femaleCHOAgeStand, echo=FALSE}
choDeathData %>% 
  filter(sex == "Female") %>% 
  ggplot(aes(x=choAgeStandValue, y=totalCVDvalue)) +
  geom_point(shape=1) +    # Use hollow circles
  geom_smooth(method = lm) +
  theme_light(9) +
  labs(x = "Cholesterol Mean mmol/L\n", y = "Female CVD Death rate per 100,000", 
       linetype = NULL, color = NULL,
       title = "Female Cardiovascualr Death Rate vs. Mean Cholesteral (Age Standardized) (by region)\n")

```
