---
title: "World Health Organization (WHO) Cholesterol Data vs. Death Rates"
author: "Steve Bargelt"
date: "May 5, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Started

### Load packages

Package loading:
```{r load-packages, message=FALSE}
library(WHO)
library(plyr)
library(dplyr)
library(ggplot2)
library(scales)
library(showtext)
library(scatterD3)
library(stringr)
library(RColorBrewer)
library(dygraphs)
```

### Data

Our investigation will focus on World Health Organization data on cholesterol data and correlation to non-communicable diseases as casue of death. 

Load the data file... you can find a description of my file preparation and ALL source code at [Github](https://github.com/stevebargelt/WHO-Data "WHO Cholesterol Data")

```{r load-data}

load(file="choDeathData.RData")

```

## Age Standardized vs. Crude Cholesterol Numbers

Must decide on one or the other since not all values are present for all data points. Decided to use Age Standardized value since there are ~20 more complete cases.

Mean Age Standardized: `r mean(choDeathData$choAgeStandValue, na.rm = TRUE)` <br />
Mean Crude: `r mean(choDeathData$choCrudeValue, na.rm = TRUE)` <br />
Median Age Standardized: `r median(choDeathData$choAgeStandValue, na.rm = TRUE)` <br />
Median Crude: `r median(choDeathData$choCrudeValue, na.rm = TRUE)` <br />
IQR Age Standardized: `r IQR(choDeathData$choAgeStandValue, na.rm = TRUE)` <br />
IQR Crude: `r IQR(choDeathData$choCrudeValue, na.rm = TRUE)` <br />
Number of complete samples Crude: `r  nrow(choDeathData[complete.cases(choDeathData[,c(7,8)]),])` <br />
Number of complete samples Age Standardized: `r nrow(choDeathData[complete.cases(choDeathData[,c(7,9)]),])` <br />

```{r remove choAgeStanValue}
#choDeathData$choCrudeValue <- NULL
choDeathData <- choDeathData[complete.cases(choDeathData[,c(7,8)]),]

```

## Including Plots

You can also embed plots, for example:

```{r maleCHOCrude, echo=FALSE}
r2<-ddply(choDeathData,.(sex),function(x) summary(lm(x$totalCVDvalue ~ x$choAgeStandValue))$r.squared)
names(r2)<-c("sex","r2")

r<-ddply(choDeathData,.(sex),function(x) cor(x$totalCVDvalue, x$choAgeStandValue, , method="pearson"))
names(r)<-c("sex","r")

g<-ggplot(choDeathData, aes(x=choAgeStandValue,y=totalCVDvalue,group=sex))
g<-g+geom_point(aes(color=sex))
g<-g+geom_text(data=r2,aes(color=sex, label = paste("R^2: ", r2,sep="")),parse=T,x=5.5,y=c(800,750), show.legend=TRUE)
g<-g+geom_text(data=r,aes(color=sex, label = paste("R: ", r,sep="")),parse=T,x=5.5,y=c(700,650), show.legend=TRUE)
g<-g+theme_light(9)
g+labs(x = "Age standardized cholesterol Mean mmol/L\n", y = "CVD Death rate per 100,000", 
       linetype = NULL, color = NULL,
       title = "Cardiovascualr Death Rate vs. Mean Cholesteral (Age Standardized) (by region)\n")


scatterD3(x = choDeathData$choAgeStandValue, 
          y = choDeathData$totalCVDvalue, 
          lab = "", #choDeathData$country,
          #size_var = comb$InfantMortality,
          col_var=choDeathData$sex,
          symbol_var = choDeathData$region,
          xlab = "Cholesterol mmol//L", 
          ylab = "Death rate per 100,000 from all CVD", 
          col_lab = "Gender",
          symbol_lab = "Regions",
          size_lab = "Size Label",
          lasso = TRUE, 
          ellipses = TRUE)
          

choDeathData %>% 
  filter(sex == "Male") %>% 
  ggplot(aes(x=choCrudeValue, y=totalCVDvalue)) +
  geom_point(shape=1) +    # Use hollow circles
  geom_smooth(method = lm) +
  theme_light(9) +
  labs(x = "Cholesterol Mean mmol/L\n", y = "Male CVD Death rate per 100,000", 
       linetype = NULL, color = NULL,
       title = "Cardiovascualr Death Rate vs. Mean Cholesteral (crude) (by region)\n")

```

```{r femaleCHOCrude, echo=FALSE}
choDeathData %>% 
  filter(sex == "Female") %>% 
  ggplot(aes(x=choCrudeValue, y=totalCVDvalue)) +
  geom_point(shape=1) +    # Use hollow circles
  geom_smooth(method = lm) +
  theme_light(9) +
  labs(x = "Cholesterol Mean mmol/L\n", y = "Female CVD Death rate per 100,000", 
       linetype = NULL, color = NULL,
       title = "Cardiovascualr Death Rate vs. Mean Cholesteral (crude) (by region)\n")

```

```{r maleCHOAgeStand, echo=FALSE}
choDeathData %>% 
  filter(sex == "Male") %>% 
  ggplot(aes(x=choAgeStandValue, y=totalCVDvalue)) +
  geom_point(shape=1) +    # Use hollow circles
  geom_smooth(method = lm) +
  theme_light(9) +
  labs(x = "Cholesterol Mean mmol/L\n", y = "Male CVD Death rate per 100,000", 
       linetype = NULL, color = NULL,
       title = "Cardiovascualr Death Rate vs. Mean Cholesteral (Age Standardized) (by region)\n")

```

```{r femaleCHOAgeStand, echo=FALSE}
choDeathData %>% 
  filter(sex == "Female") %>% 
  ggplot(aes(x=choAgeStandValue, y=totalCVDvalue)) +
  geom_point(shape=1) +    # Use hollow circles
  geom_smooth(method = lm) +
  theme_light(9) +
  labs(x = "Cholesterol Mean mmol/L\n", y = "Female CVD Death rate per 100,000", 
       linetype = NULL, color = NULL,
       title = "Cardiovascualr Death Rate vs. Mean Cholesteral (Age Standardized) (by region)\n")

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
